---
title: DATA 606 Hakim - Project
author: "Jawaid Hakim"
date: "`r Sys.Date()`"
output:
     powerpoint_presentation:
            reference_doc: data606-finalproject.potx
---

```{r load-library, include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(infer)
library(ggstatsplot)
```

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
df <- read_csv(file = "https://raw.githubusercontent.com/himalayahall/CUNY-DATA606/master/Project/resume_callback.csv",
               col_types = cols(
                   
                    job_ad_id = col_character(),
                    firstname = col_character(),
                    
                    job_city = col_factor(),
                    job_industry = col_factor(),
                    job_type = col_factor(),
                    job_ownership = col_factor(),
                    job_req_school = col_factor(),
                    resume_quality = col_factor(ordered = TRUE, levels = c('low', 'high')),
                    race = col_factor(),
                    gender = col_factor(),
                    received_callback = col_factor(),
 
                    job_req_min_experience = col_character(),
                    years_experience = col_integer(),
                    years_college = col_factor(ordered = TRUE, levels = c('0', '1', '2', '3', '4')),
                    
                    .default = col_logical()
                     
               ))
```

```{r mutate-data, include=FALSE, message=FALSE, warning=FALSE}
df$received_callback_cat <- recode_factor(df$received_callback, `0` = 'No',  `1` = 'Yes')
df <- df |> select(-job_ad_id)
```

# Introduction

-   Data comes from a study that sought to understand the influence of race (and gender) on job application callback rates

-   Researchers monitored job postings in Boston and Chicago during 2001 and 2002

-   Randomly generated resumes to send to job postings

-   Randomly assigned a *name* to the resume that would communicate the race and gender of the applicant

-   First names for the study were selected that would predominantly be recognized as belonging to **black** or **white** individuals

**Research Question**: Does the perceived **race** of applicants have a meaningful impact on the **callback rate**?

# Data

- [OpenIntro - Which resume attributes drive job callbacks?](https://www.openintro.org/data/index.php?data=resume)
- This was an **experiment** conducted over several months during 2001 and 2002 in Boston and Chicago
- Each of the cases (4870) represents a randomly generated resume
- Categorical response variable is $received\_callback$
- Categorical explanatory variable is $race$

# Callback Proportions - Group by Callback Rate

```{r echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

grouped_ggbarstats(
  data         = df,
  x            = race,
  y            = received_callback_cat,
  grouping.var = received_callback_cat,
  package      = "wesanderson",
  palette      = "Darjeeling2",
  label = 'both'
)
```

# Callback Proportions - Group by Race

```{r echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

grouped_ggbarstats(
  data         = df,
  x            = received_callback_cat,
  y            = race,
  grouping.var = race,
  package      = "wesanderson",
  palette      = "Darjeeling2",
  label = 'both'
)
```
```{r set-seed, include=FALSE, message=FALSE, warning=FALSE}
set.seed(1234)
```

# Calculate sample proportion

```{r compute-sample-props, message=FALSE, warning=FALSE}
sample_props <- df |>
                    rep_sample_n(size = 500, 
                                 reps = 1000, 
                                 replace = TRUE) |>
                    count(received_callback_cat) |>
                    mutate(p_hat = n /sum(n)) |>
                    filter(received_callback_cat == "Yes")
```

# Sample proportion

```{r plot-sample-props, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Sample Prop - Received Callback"}
ggplot(data = sample_props, aes(x = p_hat)) +
  geom_histogram(binwidth = 0.005)

```

# Sample Proportion - 95% Confidence Interval

```{r compute-ci, message=FALSE, warning=FALSE}
ci <- df |>
  specify(response = received_callback_cat, success = 'Yes') |>
  generate(reps = 1000, type = 'bootstrap') |>
  calculate(stat = 'prop') |>
  get_ci(level = 0.95)
print(ci)
```

# Independence hypothesis test

$H_0$: assumes there is no association between callback rate and the perceived race of the candidate

$H_A$: assumes there is an association between the callback rate and perceived race of the candidate

# Chi-square Testing framework

Assumptions for using Chi-square:

1.  Both variables are categorical
2.  Observations are independent
3.  Individuals can only belong to one cell in the contingency table. That is, cells in the table are mutually exclusive -- an individual cannot belong to more than one cell
4.  There must be at least 5 frequencies in each cell

Our dataset satisfies all above assumptions.

# Observed Chi-square

```{r message=FALSE, warning=FALSE}
obs_indep_stat <- df |>
    specify(received_callback_cat ~ race, 
            success = 'Yes') |>
    hypothesize(null = 'independence') |>
    calculate(stat = 'Chisq', 
              c('white', 'black'))
```

```{r echo=FALSE}
# ggbarstats(
#   data = df,
#   x = received_callback_cat,
#   y = race,
# grouping.var = race
# ) +
#   labs(caption = NULL) # remove caption
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
obs_indep_stat$stat
```

# Simulated null distribution

The null distribution is what we would expect to see if the response and explanatory variables were truly independent.

```{r message=FALSE, warning=FALSE}
sim_null_dist <- df |>
  specify(received_callback_cat ~ race, 
          success = 'Yes') |>
  hypothesize(null = 'independence') |>
  generate(reps = 500, 
           type = 'permute') |>
  calculate(stat = 'Chisq', 
            c('white', 'black'))
```

# Null distribution plot with observed stat

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
sim_null_dist |>
  visualize(method = 'both') +  
  shade_p_value(obs_stat = obs_indep_stat,
                direction = NULL)
```

Plot of the null distribution shows the observed test statistic would be unlikely if there was no association between $race$ and $callback\_rate$. Consequently, we reject $H_0$ in favor $H_A$.

# Other Variables

This study looked at $race$ as the explanatory variable. However, the dataset contains other variables (see below) that may be relevant. A multiple regression model could be used to identify combinations of explanatory variables.

1.  Gender
2.  Resume Quality
3.  Computer Skills
4.  Volunteer
5.  ...

# Generalized linear model

``` {}
                    Estimate Std. Error z value Pr(>|z|)    
(Intercept)         -1.8120     0.1499 -12.091  < 2e-16 ***
genderm             -0.1919     0.1334  -1.439    0.150    
resume_quality.L     0.3022     0.1157   2.612    0.009 ** 
computer_skillsTRUE -0.3787     0.1378  -2.748    0.006 ** 
volunteerTRUE       -0.2168     0.1617  -1.340    0.180    
raceblack           -0.4307     0.1075  -4.006 6.18e-05 ***
```

```{r include=FALSE, message=FALSE, warning=FALSE}
mlm <- glm(received_callback ~ gender + resume_quality + computer_skills + volunteer +race, data = df, family = binomial(link = 'logit'))
summary(mlm)
```

# Limitations

This study was originally conducted in 2001 and 2002 in Chicago and Boston. Over the last 20 years there has been increasing awareness of bias in hiring practices. The data needs to be refreshed and more cities should be included in the study to eliminate potential regional biases.

# Conclusion

Many biases have been identified in the job market over the years. As a result, HR professionals recommend not including age, religion, nationality, sexual orientation, etc. in job applications.

This study identified the applicant's first name as a potential source of bias (!), which should raise red flags for HR professionals, job applicants, and hiring managers.

One hopes such studies will lead to greater awareness regarding subtle biases in hiring practices and the need for training and due diligence to overcome such biases.
