---
title: "Multiple linear regression"
author: ""
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

## Grading the professor

Many college courses conclude by giving students the opportunity to
evaluate the course and the instructor anonymously. However, the use of
these student evaluations as an indicator of course quality and teaching
effectiveness is often criticized because these measures may reflect the
influence of non-teaching related characteristics, such as the physical
appearance of the instructor. The article titled, "Beauty in the
classroom: instructors' pulchritude and putative pedagogical
productivity" by Hamermesh and Parker found that instructors who are
viewed to be better looking receive higher instructional ratings.

Here, you will analyze the data from this study in order to learn what
goes into a positive professor evaluation.

## Getting Started

### Load packages

In this lab, you will explore and visualize the data using the
**tidyverse** suite of packages. The data can be found in the companion
package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(GGally)
```

This is the first time we're using the `GGally` package. You will be
using the `ggpairs` function from this package later in the lab.

### The data

The data were gathered from end of semester student evaluations for a
large sample of professors from the University of Texas at Austin. In
addition, six students rated the professors' physical appearance. The
result is a data frame where each row contains a different course and
columns represent variables about the courses and professors. It's
called `evals`.

```{r}
glimpse(evals)
```

We have observations on 21 different variables, some categorical and
some numerical. The meaning of each variable can be found by bringing up
the help file:

```{r help-evals, eval=FALSE}
?evals
```

## Exploring the data

1.  Is this an observational study or an experiment? The original
    research question posed in the paper is whether beauty leads
    directly to the differences in course evaluations. Given the study
    design, is it possible to answer this question as it is phrased? If
    not, rephrase the question.

This is an observational study, there is no control group.

Beauty is in the eye of the beholder (highly subjective) and there is no
control group against which to compare results. In addition, only six
students rated the professors on their physical appearance so the sample
size is small.

Therefore, it is not possible to answer the question as posed because
there may be other factors that contribute to teacher evaluations apart
from a teacher's perceived attractiveness. Perhaps a better phrasing
might be 'whether there is a relationship between a teacher's perceived
attractiveness and course evaluations'.

2.  Describe the distribution of `score`. Is the distribution skewed?
    What does that tell you about how students rate courses? Is this
    what you expected to see? Why, or why not?

```{r}
hist(evals$score, xlab = 'Score')
```

The distribution for `score` is left-skewed, meaning more students rated
their professors more positively. This is somewhat surprising as one
would expect the distribution to be normal. The skew might be due to
other factors such as the selected departments, course levels, etc.

Although this seems to be true regardless of `rank`, `teaching`
professors have a shorter left-tail (teaching professors were, in
general, rated higher than tenured and tenure track professors).

```{r}
ggplot(data = evals, aes(x = score, fill = rank)) + geom_histogram()
```

3.  Excluding `score`, select two other variables and describe their
    relationship with each other using an appropriate visualization.

```{r}
boxplot(evals$bty_avg ~ evals$pic_color, main = "Boxplot of Beauty Average Score by Pic Color", ylab = "Beauty Average", xlab = "Pic Color")
```

Teachers with Black&White photos received higher Beauty Average scores
(3-8) than those with Color photos (1.5-5).

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking
teachers are evaluated more favorably. Let's create a scatterplot to see
if this appears to be the case:

```{r scatter-score-bty_avg}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()
```

Before you draw conclusions about the trend, compare the number of
observations in the data frame with the approximate number of points on
the scatterplot. Is anything awry?

There are 463 observations and the scatterplot shows far fewer.

4.  Replot the scatterplot, but this time use `geom_jitter` as your
    layer. What was misleading about the initial scatterplot?

```{r scatter-score-bty_avg-jitter}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter()
```

There was overplotting in the initial scatterplot - i.e. multiple points
mapped to same coordinates.

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict
    average professor score by average beauty rating. Write out the
    equation for the linear model and interpret the slope. Is average
    beauty score a statistically significant predictor? Does it appear
    to be a practically significant predictor?

Add the line of the bet fit model to your plot using the following:

```{r scatter-score-bty_avg-line-se}
lm_bty <- lm(evals$score ~ evals$bty_avg)
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm")
summary(lm_bty)
```

The blue line is the model. The shaded gray area around the line tells
you about the variability you might expect in your predictions. To turn
that off, use `se = FALSE`.

```{r scatter-score-bty_avg-line}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)
```

6.  Use residual plots to evaluate whether the conditions of least
    squares regression are reasonable. Provide plots and comments for
    each one (see the Simple Regression Lab for a reminder of how to
    make these).

```{r}
score_bty_avg = lm(evals$score ~ evals$bty_avg)
ggplot(data = score_bty_avg, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```

Fitted values are evenly distributed above and below the zero line. This
shows a **liner relationship** between the two variables.

```{r}
ggplot(data = score_bty_avg, aes(sample = .resid)) +
  stat_qq()
```

Residuals appear to be **nearly normal** with a left-skew.

## Multiple linear regression

The data set contains several variables on the beauty score of the
professor: individual ratings from each of the six students who were
asked to score the physical appearance of the professors and the average
of these six scores. Let's take a look at the relationship between one
of these scores and the average beauty score.

```{r bty-rel}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_point()

evals %>% 
  summarise(cor(bty_avg, bty_f1lower))
```

As expected, the relationship is quite strong---after all, the average
score is calculated using the individual scores. You can actually look
at the relationships between all beauty variables (columns 13 through
19) using the following command:

```{r bty-rels}
evals %>%
  select(contains("bty")) %>%
  ggpairs()
```

These variables are collinear (correlated), and adding more than one of
these variables to the model would not add much value to the model. In
this application and with these highly-correlated predictors, it is
reasonable to use the average beauty score as the single representative
of these variables.

In order to see if beauty is still a significant predictor of professor
score after you've accounted for the professor's gender, you can add the
gender term into the model.

```{r scatter-score-bty_avg_pic-color}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the
    conditions for this model are reasonable using diagnostic plots.

```{r}
ggplot(data = m_bty_gen, aes(sample = .resid)) +
  stat_qq() + stat_qq_line()
```

Although the points drift off below the normal line towards the upper
end, most of the data falls along the normal line.

```{r}
plot(m_bty_gen$residuals)
abline(h = 0, lty = 3)
```

The residuals are spread evenly along the zero line. Conditions for this
model have been met.

8.  Is `bty_avg` still a significant predictor of `score`? Has the
    addition of `gender` to the model changed the parameter estimate for
    `bty_avg`?

```{r}
summary(m_bty_gen)
```

`bty_avg` is still a significant predictor of score and adding the
gender variable to the model has increased the poarameter estimate from
0.06664 to 0.07416. However, the R-square is very low for this model.

Note that the estimate for `gender` is now called `gendermale`. You'll
see this name change whenever you introduce a categorical variable. The
reason is that R recodes `gender` from having the values of `male` and
`female` to being an indicator variable called `gendermale` that takes a
value of $0$ for female professors and a value of $1$ for male
professors. (Such variables are often referred to as "dummy" variables.)

As a result, for female professors, the parameter estimate is multiplied
by zero, leaving the intercept and slope form familiar from simple
regression.

$$
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
$$

```{=html}
<!-- We can plot this line and the line corresponding to those with color pictures
with the following  -->
```
<!-- custom function. -->

```{r twoLines}
ggplot(data = evals, aes(x = bty_avg, y = score, color = pic_color)) +
 geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

9.  What is the equation of the line corresponding to those with color
    pictures? (*Hint:* For those with color pictures, the parameter
    estimate is multiplied by 1.) For two professors who received the
    same beauty rating, which color picture tends to have the higher
    course evaluation score?

$$
\begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (1) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2
\end{aligned}
$$

The decision to call the indicator variable `gendermale` instead of
`genderfemale` has no deeper meaning. R simply codes the category that
comes first alphabetically as a $0$. (You can change the reference level
of a categorical variable, which is the level that is coded as a 0,
using the`relevel()` function. Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and
    `rank` added in. How does R appear to handle categorical variables
    that have more than two levels? Note that the rank variable has
    three levels: `teaching`, `tenure track`, `tenured`.

```{r}
m_bty_rank = lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```

Since rank variable has three levels (teaching, tenure track and
tenured), R has added another line into the regression summary to
account for it. R leaves out one level but mentions the rest as
variables.

The interpretation of the coefficients in multiple regression is
slightly different from that of simple regression. The estimate for
`bty_avg` reflects how much higher a group of professors is expected to
score if they have a beauty rating that is one point higher *while
holding all other variables constant*. In this case, that translates
into considering only professors of the same rank with `bty_avg` scores
that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on
rank, gender, ethnicity, language of the university where they got their
degree, age, proportion of students that filled out evaluations, class
size, course level, number of professors, number of credits, average
beauty rating, outfit, and picture color.

11. Which variable would you expect to have the highest p-value in this
    model? Why? *Hint:* Think about which variable would you expect to
    not have any association with the professor score.

I would expect the number of credits to not have any association with
the professor score.

Let's run the model...

```{r m_full, tidy = FALSE}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```


12. Check your suspicions from the previous exercise. Include the
model output in your response.

My guess was off the mark. In fact the variable `cls_profsingle` (number
of professors teaching the course: single, multiple) has the highest
p-value in this model.

13. Interpret the coefficient associated with the ethnicity variable.

The `ethnicity` variable increases the `score` by 0.1234929, all other
variables being held constant.

14. Drop the variable with the highest p-value and re-fit the model. Did
    the coefficients and significance of the other explanatory variables
    change? (One of the things that makes multiple regression
    interesting is that coefficient estimates depend on the other
    variables that are included in the model.) If not, what does this
    say about whether or not the dropped variable was collinear with the
    other explanatory variables?

```{r}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level +  cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

After dropping `cls_profsingle` the coefficients and significance of the
other explanatory variables has changed very slightly. My guess is that
the `cls_profsingle` was collinear with the other explanatory variables,
and thus dropping it did not change the model significantly.

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the
    linear model for predicting score based on the final model you
    settle on.

```{r}
lm_backwards = lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval + cls_credits + bty_avg + pic_color, data = evals)
summary(lm_backwards)
```

$$
\begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times rank + \hat{\beta}_2 \times ethnicity + \hat{\beta}_3 \times gender + \hat{\beta}_4 \times language \times \hat{\beta}_5 \times age + \hat{\beta}_6 \times cls\_perc\_eval + \hat {\beta}_7 \times cls\_credits + \hat{\beta}_8 \times bty\_avg + \hat{\beta}_9 \times pic\_color
\end{aligned}
$$ 

16. Verify that the conditions for this model are reasonable using
    diagnostic plots.

```{r}
ggplot(data = lm_backwards, aes(sample = .resid)) +
  stat_qq() + stat_qq_line()
```

```{r}
plot(lm_backwards$residuals)
abline(h = 0, lty = 3)
```

17. The original paper describes how these data were gathered by taking
    a sample of professors from the University of Texas at Austin and
    including all courses that they have taught. Considering that each
    row represents a course, could this new information have an impact
    on any of the conditions of linear regression?

Clearly the independence assumption is not longer valid. 

18. Based on your final model, describe the characteristics of a
    professor and course at University of Texas at Austin that would be
    associated with a high evaluation score.

A high evaluation scoure at UT-Austin would be associated with a professor with a *teaching* rank, non-minority, male, English educated, with a Black&White profile photograph.

19. Would you be comfortable generalizing your conclusions to apply to
    professors generally (at any university)? Why or why not?

I would not be comfortable generalizing my conclusions to professors generally. In order to make a generalized conclusion I think it would be necessary to take a much larger sample of students from across a broader set of universities (public/private, 2/4 year colleges, etc) reflecting a large cross-section of student population.

------------------------------------------------------------------------
